Predictive modeling using genomic annotations

### 2018-11-26 Manuscript/manuscript.Rmd
manuscript edits

@@

Careful consideration must also be taken when deciding which performance metric(s) to present when evaluating predictive models. A common metric to consider is the the area under the ROC curve (AUC), as well as others including accuracy, sensitivity, and specificity. The main argument in favour of ROC curves is that they do not depend on a threshold, whose selection is thought to be subjective. However, when the prevalence of an event is low--that is, a genomic bin containing a TAD boundary)--mean probabilities are biased towards the most common event (Jorge M. Lobo). Likewise, metrics like accuracy and sensitivity and specificity do not incorporate the entirety of the confusion matrix created when validating predictive models on a test data set. Therefore, each of these metrics are likely to be inflated when evaluating model performance on imbalanced data. More robust metrics like the F1-score and Matthew's Correlation Coefficient are more suitable for this type of analysis and can highlight problems associated with applying predictive models on imbalanced data.

@@

Publicly available topologically associating domain data for the GM12878 cell line was obtained using the arrowhead algorithm from the Juicer toolbox (Rao et. al). The domain data was constructed from in situ Hi-C contact matrices at 5kb, 25kb, 50kb, and 100kb resolutions. Identified TAD boundaries were represented by their location in the genome (hg19 human genome assembly), including chromosome, start and end coordinates for chromosomes 1 through 22. The start and end coordinates were concatenated, sorted, and unique coordinates representing the borders demarcating TADs were obtained.

@@

Annotation data, in the form of functional genomic elements from ChIP-seq experiments were obtained from the Encyclopedia of DNA Elements (ENCODE) Consortium. The annotations consisted of a variety of different chromatin states, histone marks, and DNase I hypersensitive sites. Each annotation was represented by their location throughout the genome via chromosomal coordinates. An example of a specific annotation, which is labeled as a functional genomic element, is presented in Figure 1. It is represented by the green segments that overlap with the genomic bins.

Using each genomic element, and their relationship with the genomic bins, we created three types of predictors that were used as features in downstream modelling. Figure 2 presents an illustration of each type of predictor that was considered. The predictors consisted of two different types of overlaps, and a distance type and are described in detail below:

   1.	Overlap Counts (OC): For each genomic bin, the total number of overlapped regions between a functional genomic element and a bin was calculated. This was defined as the overlap counts (OC) and is illustrated in the left part of Figure 2. From the first bin, in the highlighted region, we see that there are two regions defined by the particular functional genomic element that overlap within the same bin. Therefore the OC for that bin, given that feature, was determined to be 2. Bins that did not overlap with a feature were assigned a value of zero.

   2.	Overlap Percent (OP): For each genomic bin, the percent overlap between regions defined by a particular functional element and a bin was calculated. From the middle region of Figure 2, the feature width was defined as the number of base pairs that overlapped with the genomic bin. The percent was then calculated by dividing the feature width with the bin width. If multiple overlaps existed, the feature width was defined as the sum of the total number of base pairs that overlapped with a particular bin. Note that the bin width remained constant for a given resolution (either 5kb, 25kb, 50kb, or 100kb). Similar to the OC predictor type, bins with no overlaps were given to be zero.

   3.	Distance: For each genomic bin, the distance (in base pairs) from the center of the bin to the center of the nearest region of a respective functional element was calculated. The distance predictors were then normalized using a log base 2 transformation prior to modelling.

@@

After choosing one of the aforementioned data resolutions, predictor types, and variable reduction techniques, we then evaluated three re-sampling techniques used to create balanced classes.  For each resampling technique, the majority class is defined as the set of genomic bins that did not include a TAD boundary, while the minority class is defined as the set of bins that included a TAD boundary. The following details the re-sampling techniques considered in this paper:

>>>>

[Mention Feature Engineering]

@@

Careful consideration must also be taken when deciding which performance metric(s) to present when evaluating predictive models. A common metric to consider is the the area under the ROC curve (AUC), as well as others including accuracy, sensitivity, and specificity.

The popularity surrounding the use of ROC curves is that they do not depend on a threshold. However, when the prevalence of an event is low--that is, a genomic bin containing a TAD boundary--mean probabilities are biased towards the majority class. Likewise, metrics like accuracy and sensitivity and specificity do not incorporate the entirety of the confusion matrix created when validating predictive models on a test data set. Therefore, each of these metrics are likely to be inflated when evaluating model performance on imbalanced data. More robust metrics like the F1-score and Matthew's Correlation Coefficient are more suitable for this type of analysis and can highlight problems associated with applying predictive models on imbalanced data.

[Mention Previous Methods]

@@

Topologically associating domain data for the GM12878 cell line was obtained using the arrowhead algorithm from the Juicer toolbox (https://github.com/aidenlab/juicer). Publically available domain data was obtained from in situ Hi-C contact matrices at 5kb resolution [-@rao20143d]. Additional domain data was obtained from Dali et al. at 25kb, 50kb, and 100kb resolutions [-@dali2017critical].

Identified TAD boundaries were represented by their location in the genome (hg19 human genome assembly), including chromosome, start and end coordinates for chromosomes 1 through 22. The start and end coordinates were concatenated, sorted, and unique coordinates representing the borders demarcating TADs were obtained.

@@

Annotation data, in the form of functional genomic elements from ChIP-seq experiments were obtained from the Encyclopedia of DNA Elements (ENCODE) Consortium. The annotations consisted of a variety of different chromatin states, histone marks, and DNase I hypersensitive sites. Table 1 presents the total list of genomic  annotations considered in this study and the sources of downloadable files. Each annotation was represented by their location throughout the genome via chromosomal coordinates. An example of a specific annotation, which is labeled as a functional genomic element, is presented in Figure 1. It is represented by the green segments that overlap with the genomic bins.

##Feature Engineering

Using each genomic element, we created three types of predictors that were used as features to analyze their relationships with TAD boundaries in downstream modelling. Figure 2 presents an illustration of each type of predictor that was considered. The predictors consisted of two different types of overlaps, and a distance type. They are described in detail below:

   1.	Overlap Counts (OC): For each genomic bin, the total number of overlapped regions between a functional genomic element and a bin was calculated. This was defined as the overlap counts (OC) and is illustrated in the left part of Figure 2. Within each bin (marked by the blue dashed lines), the total number of overlapped regions of the genomic element (represented by the horizontal green segments) are tallied and marked below. In the case of no overlaps the value is given as 0.

   2.	Overlap Percent (OP): For each genomic bin, the percent overlap between regions defined by a particular functional element and a bin was calculated. From the middle diagram of Figure 2, the feature width was defined as the number of base pairs that overlapped with the genomic bin. The percent was then calculated by dividing the feature width with the bin width. If multiple overlaps existed, the feature width was defined as the sum of the total number of base pairs that overlapped with a particular bin. Note that the bin width remained constant for a given resolution (either 5kb, 25kb, 50kb, or 100kb). Similar to the OC predictor type, bins with no overlaps were given to be 0.

   3.	Distance: For each genomic bin, the distance (in base pairs) from the center of the bin to the center of the nearest region of a respective functional element was calculated. The right most diagram of Figure 2 depicts an example of how distance was calculated. The vertical green segments represent the centers of the respective genomic element. Each distance, $D_{i}$, represented by the blue brackets are the number of base pairs from the center of the bins to the center of the nearest genomic elemental region.

[Mention transformations used]

##Variable Reduction

We evaluated the use of two variable reduction algorithms, the LASSO and Elastic-Net regularizations. Each algorithm was performed prior to re-sampling. For the LASSO, the $\lambda$ penalization term was tuned over a grid of 10 values on an exponential scale ranging from 0.01 to 10. The additional $\alpha$ term for the Elastic-Net was tuned

@@

Figure 3 details the ensemble framework proposed and evaluated in this study. For each resolution and predictor type chosen, the data, composed of majority and minority classes, was split into a 7:3 ratio of training to testing sets. For the training set, a variable reduction technique was performed. The feature space was reduced and a resampling technique was applied. A classification algorithm in the form of a random forest was performed and validated using the test set. Details of each step in the framework are provided below.

###Training and Testing Sets

###Variable Reduction

###Re-Sampling Techniques

After choosing one of the aforementioned data resolutions, predictor types, and variable reduction algorithms, we then evaluated three re-sampling techniques used to create balanced classes.  For each resampling technique, the majority class is defined as the set of genomic bins that did not include a TAD boundary, while the minority class is defined as the set of bins that included a TAD boundary. The following details the re-sampling techniques considered in this paper:
